# Reproducible Research: Peer Assessment 1
This data set includes information about step activitiy, binned into 5 minute intervals, for a single person over 61 days (October - November of 2012). 

## Loading and preprocessing the data

```{r, echo=TRUE}
data <- read.csv("activity.csv")
data$date_as_obj <- as.Date(data$date, "%Y-%m-%d")
data$hour  <- data$interval %/% 100
data$minute  <- data$interval %% 100
data$date_time <- strptime(paste(data$date, " ", data$hour, ":", data$minute, sep = ""), format("%Y-%m-%d %H:%M"))

```

## What is mean total number of steps taken per day?

Histogram is the frequency of total steps in a day binned into 15 segments.

```{r, echo=TRUE}
total_steps_in_a_day  <-  tapply(data[!is.na(data$steps),]$steps, data[!is.na(data$steps),]$date_as_obj, sum)
hist(total_steps_in_a_day, breaks = 15, xlab = "Number of Steps Per Day", main = "Daily Step Frequency (Oct - Nov)")

mean_steps <- mean(tapply(data[!is.na(data$steps),]$steps, data[!is.na(data$steps),]$date_as_obj, sum))
median_steps <- median(tapply(data[!is.na(data$steps),]$steps, data[!is.na(data$steps),]$date_as_obj, sum))

```

The mean number of steps in a day is -- STRING INTERPOLATE -- and the median number of steps in a day is --STRING INTERPOLATE--.


## What is the average daily activity pattern?

```{r, echo=TRUE}
    data$day_time <- data$date_time - trunc(data$date_time, "days")
    data_na_removed <- data[!is.na(data$steps), ]
average_steps_throughout_day <- tapply(data_na_removed$steps, data_na_removed$date_time - trunc(data_na_removed$date_time, "days"), mean)
    average_steps_throughout_day <- ddply(.data = data_na_removed, .(day_time), summarize, ave_steps = mean(steps))    
    plot(strptime(as.character(seconds_to_period(average_steps_throughout_day$day_time)), format="%HH %MM %SS"), average_steps_throughout_day$ave_steps, type = "l", xlab = "Time of Day", ylab = "Average Number of Steps")
    
```

interval with max average steps is the 5 minute period at 08:35 with an average of 206.1698 steps

```{r, echo=TRUE}
    period_with_max_ave_steps <- average_steps_throughout_day[average_steps_throughout_day$ave_steps == max(average_steps_throughout_day$ave_steps), ]$day_time
    
```


## Imputing missing values

```{r, echo=TRUE}
    data_na_removed$day_of_week  <- as.factor(weekdays(data_na_removed$date_as_obj))
    qplot(steps, data = data_na_removed, facets = day_of_week~., binwidth = 10)
    ave_day_of_week_steps <- ddply(.data = data_na_removed, .(weekdays(date_as_obj)), summarize, ave_steps = mean(steps))
```

## Are there differences in activity patterns between weekdays and weekends?
